---
title: WIP- Climate change will cook chestnut-breasted whiteface
author:
  - name: Ned L Ryan-Schofield
    orcid: 0000-0002-4997-6560
    corresponding: true
    email: ned.ryan-schofield@bushheritage.org.au
    roles:
      - Conceptualisation
      - Data Collection
      - Analysis
      - Manuscript preparation
    affiliations:
      - Bush Heritage Australia
      - Charles Darwin University
  - name: Darcy A Whittaker
    orcid:  
    corresponding: false
    roles:  
      - Conceptualisation
      - Data Collection
      - Manuscript preparation
    affiliations:
      - Bush Heritage Australia
  - name: Graeme R Finlayson
    orcid: 0000-0002-6770-2520 
    corresponding: false
    roles:  
      - Conceptualisation
      - Manuscript preparation
    affiliations:
      - Bush Heritage Australia
  - name: Lucy McCormack
    orcid:  
    corresponding: false
    roles:  
      - Analysis
      - Manuscript preparation
    affiliations:
      - Bush Heritage Australia
  - name: Patrick L Taggart
    orcid: 0000-0001-9523-0463 
    corresponding: false
    roles:  
      - Conceptualisation
      - Data Collection
      - Analysis
      - Manuscript preparation
    affiliations:
      - Bush Heritage Australia
keywords:
  - Climate change
  - Heatwaves
abstract: |
  tba ...
plain-language-summary: |
  tba ...
key-points:
  - tba
date: last-modified
bibliography: references.bib
citation:
  container-title: Earth and Space Science
number-sections: true
---

NRS NOTES 28/5/26: - Figures are produced by the scripts in https://github.com/nedschofield/cbwf\_-manuscript_project.git, stored locally, and just called from this manuscript. - After I finalise the analysis I'll port it all over into notebooks/ so that it is all reproducible, but for now I want to keep just one version of the analysis.

## Introduction

With current and ongoing high rates of extinction, the importance of ecological monitoring is fundamental to understanding the natural environment and conserving species [@lindenmayer2018]. Identifying declining populations through effective biodiversity monitoring can prove critical in implementing timely conservation measures [@baker2019][@butchart2006]. This is particularly pertinent for threatened, rare and range-restricted species. Such species are at greater risk of rapid and unrecognised decline due to their increased susceptibility to stochastic events [@smith2020].  

Relative to other land types, Australian arid landscapes are typically characterised by low human population densities. A marked spatial unevenness in biodiversity data is seen across Australia, with sparsely populated areas often lacking in biological records [@woinarski2021]. In theory, these factors could contribute to species declines going unnoticed in remote areas [@woinarski2021]. Arid lands contain unique biodiversity components, the loss of which can have profound impacts on the overall ecosystem [@zhang2023]. Threatened species in such environments therefore require carefully designed monitoring programs to inform conservation management and better detect population changes.  

Arid and semi-arid rangelands cover a significant proportion of Australia, and like elsewhere they are characterised by low human population density, high numbers of threatened species and high rates of extinctions [@morton1995][@woinarski2003]. Since European settlement in Australia, 20 arid land species are known to have become extinct, with numerous others currently threatened with extinction [@woinarski2019]. The arid rangelands have experienced rapid and extensive change in the time since European colonisation. While significant swathes of land still fall under Indigenous management [@renwick2017], pastoralism is one of the major land uses in the Australian arid rangelands [@letnic2006]. Pastoral activities are known to have direct and indirect impacts on arid land ecosystem function [@letnic2007]. Along with the impacts of pastoralism, other threats pose major challenges to species living in the arid rangelands, such as introduced species and climate change [@garnett2024]. Naturally boom and bust systems, arid lands experience large fluctuations in available resources and environmental conditions over time [@letnic2006]. Species living in arid environments are adapted to these changing conditions, but survival under future climate change is likely to become increasingly challenging [@conradie2020]. By the end of the 21st century, Zebra Finches (*Taeniopygia castanotis*), a small arid-adapted passerine, are expected in some regions to experience a more than threefold increase in the number of days per year that they are exposed to lethal dehydration risk [@conradie2020]. 

The Chestnut-breasted Whiteface (*Aphelocephala pectoralis*) is a small passerine of arid, stony, low shrublands in northern South Australia. A South Australian endemic, *A. pectoralis* has a restricted range and has anecdotally faced declines in some areas. Prior to 1990, the species was known from only 33 localities (5 unconfirmed) [@pedler1992]. Proving elusive through the late 19th and mid-20th century, *A. pectoralis* went undetected for long periods of time on more than one occasion (43 years between the type specimen’s discovery in 1871 and the next recorded sighting in 1914, and again for another 43 years between 1923 and 1966) [@pedler1992]. In 1990, Lynn Pedler revisited almost all the then known *A. pectoralis* localities and also discovered several new occupied sites. Using the results from these surveys, and observed population densities from a study site in the eastern part of the species’ range (Mount Lyndhurst), an overall population of 6000 individuals was estimated (exact methodology not stated) [@pedler1992]. Some populations were again resurveyed in 1999 [@pedler2000] and the Mount Lyndhurst population monitored again in 2007 [@pedler2007].  

Despite the species’ apparent rarity, restricted range and anecdotal evidence of declines in some areas, *A. pectoralis* has had little to no research or conservation attention excluding the work of Pedler et al. Listed nationally as Near Threatened and as Rare in South Australia, *A. pectoralis* lacks contemporary abundance and distribution data and its conservation status may reflect this uncertainty [@garnett2021].  

This project aimed to assess the current distribution of the Chestnut-breasted Whiteface and help inform future conservation status assessments. To achieve this, as many of the original 1990 survey sites as possible were revisited in addition to a selection of sites where the species had since been detected. Through this process, we also aimed to develop a standardised, repeatable survey protocol suitable for *A. pectoralis*. Simultaneously, we aimed to collect as much biologically relevant information on the species as possible to help to understand their ecology and inform their conservation.

## Methods {#sec-data-methods}

### Survey Region and Site Selection

```{r}
#packages
library(dplyr)

#create a path to cbwf_project location
proj.path <- "C:/Users/nedsc/Documents/git/cbwf_project/"

#survey data
d <- read.csv(file = paste0(proj.path, "data/table/CBWF_surveys_11052026.csv"))

n_sites <- length(unique(d$site))
tot_cbwf <- d |> 
  group_by(site) |>
  summarise(tot = sum(CBWF))
sites_w_cbwf <- nrow(subset(tot_cbwf, tot > 0))

```

Surveys were conducted across the known and suspected range of the Chestnut-breasted Whiteface *Aphelocephala pectoralis* in arid northern South Australia between June and September 2024. Survey sites were selected to reassess historical localities and more recent records of the species, rather than to provide a random sample of available habitat. The core set of survey sites was based on localities surveyed in [@pedler1992] during a review of the species’ distribution and status, including sites where Chestnut-breasted Whitefaces were detected and sites where surveys were unsuccessful. Additional sites were included from subsequent surveys in 1999–2000 [@pedler2000] and from more recent species records, particularly where these records represented areas not well covered by the 1990 survey network.

In total, `r n_sites` sites were surveyed. These sites were distributed across four broad regions: the Coober Pedy–Glendambo region; the Oodnadatta region, including sites between Marla and Oodnadatta and between Nilpinna and William Creek; the Marree–Leigh Creek region, including the Lyndhurst–Birdsville Track area; and the Gawler Ranges–Lake Gairdner region, including sites around Mt Ive and Nonning. Most sites occurred within arid stony plains, chenopod shrublands and associated low shrubland habitats on working pastoral properties.

Where accurate location information was available, survey cells were positioned to encompass the original historical survey locality as closely as possible. For some historical records, however, location information was imprecise or unavailable at the scale required to relocate the original survey point. In these cases, surveys were placed within the approximate locality described by the historical record and in habitat judged to be broadly comparable to that originally surveyed. A map of surveyed sites and their broad regional grouping is provided in @fig-cbwf-map.

### Survey Methods

At each site, we conducted a set of repeated area searches designed to provide standardised count data while maximising the probability of detecting a rare and potentially patchily distributed species. The standard survey unit was a 20-minute, 5-ha area search. Each site comprised ten such survey cells, giving a total surveyed area of 50 ha per site. Cells were arranged in two spatially separated transects, with five 5-ha cells in each transect. Transects were separated by 200m to reduce the risk of double-counting individuals and to minimise interference between observers and playback in adjacent transects. Each cell was surveyed once during a site visit.

Surveys were conducted on foot by a single observer per cell using binoculars. Where two observers were conducting the survey, observers surveyed separate transects independently, beginning at one end of the transect and moving sequentially through the five cells. This ensured a consistent direction of travel through each transect and reduced the likelihood of repeatedly flushing or re-counting birds between cells. In each cell, observers recorded the number of Chestnut-breasted Whiteface detected during the 20-minute search. For each Chestnut-breasted Whiteface detection, observers recorded the number of individuals detected and whether the initial detection occurred before or after call playback. Call playback was used to increase detectability. At the centre of each 5-ha cell, observers broadcast Chestnut-breasted Whiteface vocalisations consisting of 30 seconds of calls and song, followed by a 30-second listening period, then a second 30-second broadcast followed by a second 30-second listening period. The use of short, standardised playback bouts at the centre of each cell was intended to increase the probability of detecting birds present within the cell while limiting attraction from beyond the surveyed area. Playback was conducted consistently across cells so that survey effort was comparable among sites.

Surveys were generally conducted as early as practicable in the day, although the remoteness of sites and long travel distances meant that some surveys were completed later in the day. Surveys were not conducted, where avoidable, during very strong winds or other conditions likely to substantially reduce detectability. For each cell, observers recorded the date, survey start time and broad survey conditions, including wind, cloud cover, rain and temperature category. At the centre of each cell, four habitat photographs were taken facing approximately north, east, south and west. These photographs, together with field notes, were used to classify the broad habitat of each survey cell.

The ten-cell design provided replicated surveys within each site, increasing the probability of detecting Chestnut-breasted Whitefaces where present and providing repeated count data while accounting for imperfect detection. This design also balanced statistical and logistical requirements: greater replication was desirable because the species is rare and often occurs at low density, but survey effort had to remain feasible across a large, remote study region with widely separated sites.

### Data Analysis

#### Occupancy

We fitted a Bayesian single-season occupancy model in JAGS [@plummer2003] to estimate site-level occupancy probability (ψ) and detection probability (*p*). Detection probability was held constant across all sites and visits. The global model included four site-level covariates on occupancy:

- Six-month cumulative pre-survey rainfall, derived from AGCD monthly precipitation (v2-0-3) for 2024, matched to the calendar month of each site visit;
- Change in mean annual rainfall over 1990–2024 (AGCD monthly precipitation, v2-0-3), expressed as the slope of a pixel-wise linear regression on annual totals;
- Absolute change in the number of days per year with maximum temperature exceeding 40°C over 1990–2024 (AGCD daily maximum temperature, v1-0-3), calculated as the pixel-wise linear trend multiplied by the length of the time series; and
- Change in photosynthetic vegetation (PV) fractional cover, expressed as the slope of a pixel-wise linear regression on monthly values from the Guerschman et al. fractional cover product derived from MODIS (covering approximately 2001 to present).

All covariates were standardised to zero mean and unit variance before model fitting. A null model with no occupancy covariates was also fitted; the two models were compared using the Deviance Information Criterion (DIC).

All regression coefficients and intercept parameters were given weakly informative Normal(0, 1) priors on the logit scale. Models were run with 3 MCMC chains of 20,000 iterations each, with a burn-in of 5,000 and thinning by a factor of 5, yielding 9,000 retained posterior samples. Convergence was confirmed using the Gelman-Rubin diagnostic (R̂ \< 1.1 for all parameters), and model fit was assessed via a Bayesian posterior predictive check using the Freeman-Tukey statistic.

Occupancy probability was predicted spatially across the study region by applying the fitted model to 1 km² aggregated raster layers of each covariate (aggregated by a factor of 10 from the native \~100 m resolution). The September 6-month rainfall layer was used for spatial prediction, as the majority of surveys were conducted in September. Uncertainty was propagated by drawing 500 thinned samples from the joint posterior to produce 95% credible interval surfaces alongside the posterior mean.

#### Distribution modelling

To place the 2024 survey results in a historical and future climate context, we fitted a boosted regression tree (BRT) model of *A. pectoralis* occurrence probability using survey records from three time periods: approximately 1960, 1990, and 2024. The dataset combined our 2024 survey records, the 1990 survey records of @pedler1992, and historical presence records compiled by @pedler1992 (predominantly from the 1960s). All records were coded as binary presence/absence observations. For each record, period-matched climate data were extracted from AGCD gridded products: mean annual days with maximum temperature exceeding 40°C and mean annual rainfall (mm), averaged over a 10-year window centred on each survey period (1950–1960 for the 1960 period, 1980–1990 for the 1990 period, and 2014–2024 for the 2024 period; AGCD daily maximum temperature v1-0-3 and monthly precipitation v2-0-3).

The BRT model was fitted using the function *dismo::gbm.step* [@elith2008], with a Bernoulli error family, tree complexity of 3, learning rate of 0.001, and bag fraction of 0.5. The optimal number of trees was determined by cross-validation within `gbm.step`.

To project occurrence probability under future climate scenarios, we obtained NARCliM2.0 regional climate model output from the NCI THREDDS server. Bias-corrected daily maximum temperature (tasmaxAdjust) and precipitation (prAdjust) data were downloaded for five global climate models (ACCESS-ESM1-5, EC-Earth3-Veg, MPI-ESM1-2-HR, NorESM2-MM, and UKESM1-0-LL), three emissions scenarios (SSP1-2.6, SSP2-4.5, SSP3-7.0), and two WRF regional climate model configurations (R3 and R5), for two 20-year periods: mid-century (2040–2059) and late-century (2080–2099). For each GCM/SSP/RCM/period combination, we computed mean annual days exceeding 40°C and mean annual rainfall. These were then ensemble-averaged across GCMs and RCMs within each SSP×period group, producing six projected climate scenarios. The BRT model was applied to each scenario to generate gridded occurrence probability surfaces, which were reprojected and resampled to match the historical prediction grid for plotting in terra [@hijmans2026].

## Results

```{r}
#| label: fig-cbwf-map
#| echo: false
#| fig-cap: "2024 Chestnut-breasted Whiteface survey detections"

knitr::include_graphics(
  file.path(proj.path, "figures", "updated_figs", "CBWF_survey_outcomes.png")
)
```

A total of `r sum(tot_cbwf$tot)` Chestnut-breasted Whiteface were detected from `r sites_w_cbwf` of `r n_sites` sites (mean = `r mean(tot_cbwf$tot)`, sd = `r sd(tot_cbwf$tot)`). Birds were only detected at sites in the Coober Pedy-Glendambo region @fig-cbwf-map <!-- NRS: this map needs improving but will do for draft -->.

### occupancy

```{r}
#| label: load-occu-data
occu_global <- readRDS(paste0(proj.path, "outputs/occu_global.rds"))
occu_null   <- readRDS(paste0(proj.path, "outputs/occu_null.rds"))
```

Detection probability was estimated to be *p* = `r round(occu_global$mean$p, 2)` (95% CrI: `r round(occu_global$q2.5$p, 2)`–`r round(occu_global$q97.5$p, 2)`), indicating moderate detectability per survey cell. Of the `r n_sites` sites surveyed, an estimated `r round(occu_global$mean$nocc, 1)` (95% CrI: `r round(occu_global$q2.5$nocc, 0)`–`r round(occu_global$q97.5$nocc, 0)`) were inferred to be occupied, slightly less than the observed number of sites where birds were recorded (`r sites_w_cbwf`).

Of the four occupancy covariates, the absolute change in extreme heat days was the only predictor whose 95% credible interval excluded zero (@fig-coeffs), with a negative effect on occupancy probability (posterior mean ≈ −2.2 on the logit scale). Annual rainfall and PV fractional cover showed positive point estimates, and 6-month rainfall a slightly negative estimate, but all three had credible intervals that broadly overlapped zero, indicating little evidence for their effects given the data.

```{r}
#| label: fig-coeffs
#| fig-cap: "Posterior mean occupancy covariate effects with 95% credible intervals (logit scale). Points to the left of zero indicate a negative effect on occupancy probability."
#| out-width: 70%

knitr::include_graphics(
  file.path(proj.path, "figures", "updated_figs", "occu_global_covar_coeffs.png")
  )
```

The spatial prediction of occupancy probability across the study region is shown in @fig-psi-maps. The highest posterior mean ψ values are concentrated in the south-west of the species' range, centred around the Bon Bon–Glendambo area, broadly coinciding with where detections occurred in 2024. Areas in the northern and eastern parts of the historical range — where no birds were detected — have substantially lower predicted occupancy probability. The credible interval maps indicate considerable uncertainty: the lower 95% CrI surface shows that even under an optimistic estimate, high-probability areas are restricted to the Bon Bon–Glendambo core, while the upper 95% CrI suggests potentially suitable conditions across a broader area.

```{r}
#| label: fig-psi-maps
#| fig-cap: "Spatial prediction of occupancy probability ψ: posterior mean (left), lower 95% credible interval (centre), and upper 95% credible interval (right). Survey sites are shown as black points."
#| out-width: 100%
knitr::include_graphics(
  file.path(proj.path, "figures", "updated_figs", "occupancy_psi_maps.png")
  )
```

### Distribution modelling

The BRT model predicts occurrence probability across the historical study region for each of the three survey periods, where warmer colours indicate climate conditions more favourable to the species. @fig-brt-hist shows a progressive contraction of climatically suitable habitat over the past six decades. In 1960, moderate-to-high predicted occurrence probability extended across a broad area of the species' range. By 1990, and more markedly by 2024, the area of high predicted probability had contracted, with much of the northern and eastern range becoming less climatically suitable, driven predominantly by increasing extreme heat.

```{r}
#| label: fig-brt-hist
#| echo: false
#| fig-cap: "BRT-predicted occurrence probability of *A. pectoralis* across the study region for three historical survey periods (approximately 1960, 1990, and 2024), based on period-matched climate covariates (annual days exceeding 40°C and mean annual rainfall). Warmer colours indicate higher predicted occurrence probability. A progressive contraction of climatically suitable habitat is evident over the past six decades, driven predominantly by increasing extreme heat."

knitr::include_graphics(
  file.path(proj.path, "figures", "updated_figs", "brt_historical_occupancy_sat.png")
)
```

Projected occurrence probability under the NARCliM2.0 ensemble scenarios is shown in @fig-brt-fut. Across all three emissions pathways and both time periods, the model projects further reductions in climatically suitable habitat relative to present. By mid-century (2040–2059), high-probability areas are projected to be largely confined to the south-western core of the current distribution. By late century (2080–2099), suitable climate conditions are projected to contract further under all scenarios, with the most severe reductions under SSP3-7.0. Under SSP1-2.6, a modest area of suitable climate persists in the south-west through the late century period, whereas under higher emissions scenarios the majority of the species' historical range is projected to become climatically unsuitable.

```{r}
#| label: fig-brt-fut
#| echo: false
#| fig-cap: "BRT-projected occurrence probability of *A. pectoralis* under NARCliM2.0 regional climate model ensemble scenarios for mid-century (2040–2059) and late-century (2080–2099), across three emissions pathways (SSP1-2.6, SSP2-4.5, SSP3-7.0). Warmer colours indicate higher predicted occurrence probability. Climatically suitable habitat is projected to contract progressively relative to 2024 across all scenarios, with the most severe reductions under SSP3-7.0 by late century."

knitr::include_graphics(
  file.path(proj.path, "figures", "updated_figs", "brt_narclim_occupancy_sat.png")
)
```

## Discussion

This study provides the first contemporary, systematic reassessment of the distribution and status of the Chestnut-breasted Whiteface (*Aphelocephala pectoralis*) across much of its known range since the surveys of Pedler and colleagues in the 1990s and 2000s [@pedler1992; @pedler2000; @pedler2007a]. By revisiting historical localities, surveying more recent observation sites, and applying a repeatable site-based survey protocol, we provide an updated baseline for a species that has long been considered rare, poorly known and potentially declining. Overall, our results indicate that *A. pectoralis* persists in the South-West of its range,

but that its occurrence is spatially restricted and uneven. The combination of detections at some previously known or suspected sites, non-detections at others, and modelled variation in occupancy probability suggests that the species’ distribution is best understood as patchy rather than continuous across apparently suitable arid-zone landscapes.

Nevertheless, several limitations should be acknowledged. First, site selection was targeted rather than random, meaning that results are most directly applicable to known and suspected localities rather than the entire potential range. Second, the study represents a single survey period, whereas arid-zone bird occurrence may vary substantially through time. Third, some historical localities were likely imprecisely recorded, making exact resurvey difficult and complicating interpretation of apparent local absences. Fourth, remotely sensed covariates may not fully capture the fine-scale habitat structure that determines site use. Finally, although call playback can improve detection, it may also influence bird movement during surveys, and care is needed to avoid double counting or interpreting responsive movement as independent evidence of abundance.

Interestingly, there is some parts of the predicted range to the north (around Indulkana in the APY lands) which we didn't actually survey that have high probability of occupancy. Birds are frequently recorded here by bellbird tours and other birders heading up the highway.

## References

::: {#refs}
:::
