---
title: WIP- Climate change will cook chestnut-breasted whiteface
author:
  - name: Ned L Ryan-Schofield
    orcid: 0000-0002-4997-6560
    corresponding: true
    email: ned.ryan-schofield@bushheritage.org.au
    roles:
      - Conceptualisation
      - Data Collection
      - Analysis
      - Manuscript preparation
    affiliations:
      - Bush Heritage Australia
      - Charles Darwin University
  - name: Darcy A Whittaker
    orcid:  
    corresponding: false
    roles:  
      - Conceptualisation
      - Data Collection
      - Manuscript preparation
    affiliations:
      - Bush Heritage Australia
  - name: Graeme R Finlayson
    orcid: 0000-0002-6770-2520 
    corresponding: false
    roles:  
      - Conceptualisation
      - Manuscript preparation
    affiliations:
      - Bush Heritage Australia
  - name: Lucy McCormack
    orcid:  
    corresponding: false
    roles:  
      - Analysis
      - Manuscript preparation
    affiliations:
      - Bush Heritage Australia
  - name: Shawn Scott
    orcid:  
    corresponding: false
    roles:  
      - Manuscript preparation
    affiliations:
      - Bush Heritage Australia
  - name: Patrick L Taggart
    orcid: 0000-0001-9523-0463 
    corresponding: false
    roles:  
      - Conceptualisation
      - Data Collection
      - Analysis
      - Manuscript preparation
    affiliations:
      - Bush Heritage Australia
keywords:
  - Climate change
  - Heatwaves
abstract: |
  tba ...
plain-language-summary: |
  tba ...
key-points:
  - tba
date: last-modified
bibliography: references.bib
citation:
  container-title: Earth and Space Science
number-sections: true
editor:
  markdown:
    wrap: sentence
---

NRS NOTES 28/5/26:

- Figures are produced by the scripts in https://github.com/nedschofield/cbwf_project.git, stored locally, and just called from this manuscript.

- update proj.path in the first code chunk to the folder where you stored the zipped folder i sent.

- After I finalise the analysis I'll port it all over into notebooks/ so that it is all reproducible, but for now I want to keep just one version of the analysis.

- NRS 2/6: could remove all mention of collecting auxillary detection data as it wasn't used in the analysis.
  bottom of intro, and occupancy in methods.

## Introduction

Ecological monitoring is fundamental to mitigating the current and ongoing high rates of extinction and species decline [@lindenmayer2018].
Identifying declining populations through effective biodiversity monitoring can prove critical in implementing timely conservation measures [@baker2019][@butchart2006].
This is particularly pertinent for threatened, rare and range-restricted species, that are at greater risk of rapid and unrecognised decline due to their increased susceptibility to stochastic events [@smith2020].

Relative to other land types, arid landscapes are typically characterised by low human population densities.
A marked spatial unevenness in biodiversity data is seen in arid regions, with sparsely populated areas lacking biological records [@woinarski2021].
These factors likely increase the probability that species declines go unnoticed in remote areas [@woinarski2021].
Arid lands are additionally renowned for their slow recovery times due to low and irregular rainfall, shallow top soil and variable fertility, and their tendency for prolonged droughts and higher temperatures [@zhang2023].
Threatened species in such environments therefore require carefully designed monitoring programs to inform conservation management and better detect population changes.

Arid and semi-arid rangelands cover 70% of Australia's land mass [@outback].
A significant portion of Australia's arid rangelands is under Indigenous management [@renwick2017], with pastoralism otherwise being the major land use [@letnic2006].
Pastoral activities directly and indirectly impact arid rangeland ecosystem function through extensive livestock grazing and widespread pastoral infrastructure such as artificial waters and barrier or exclusion fences [@letnic2007].
Introduced species and climate change additionally pose major challenges to species occupying Australia's arid rangelands [@garnett2024].
Combined, these rapid and extensive changes have seen Australia's rangelands experience the highest rates of species extinction of any land type, including the extinction of 20 species since European colonisation [@woinarski2019].

Australia's arid rangelands are classified as boom and bust systems, whereby available resources and environmental conditions fluctuate immensely through time [@letnic2006].
Species occupying these arid rangelands have adapted to these changing conditions and resource pulses, however, survival under future climate change is likely to become increasingly challenging [@conradie2020].
Future climate predictions for Australia's rangelands estimate increased warming, longer dry spells and more intense extreme weather [@stokes2008].
For some species these changes have already been modelled.
For example, by the end of the 21st century, the Zebra Finch (*Taeniopygia castanotis*), a small arid-adapted passerine, is expected to experience more than a threefold increase in the number of days per year that they are exposed to lethal dehydration risk in some parts of its distribution [@conradie2020].

The Chestnut-breasted Whiteface (*Aphelocephala pectoralis*) is a small passerine of arid, stony, low shrublands.
They are endemic to South Australia, and are restricted to the north and north-east of the state.
Prior to 1990, the species was known from only 33 localities (5 unconfirmed) [@pedler1992].
Proving elusive through the late 19th and mid-20th century, the species went undetected for long periods of time on more than one occasion (43 years between the type specimen’s discovery in 1871 and the next recorded sighting in 1914, and again for another 43 years between 1923 and 1966) [@pedler1992].
In 1990, range-wide Chestnut-breasted Whiteface surveys revisited almost all the then known localities and also discovered several new occupied sites.
Using the results from these surveys, and observed population densities from a study site in the eastern part of the species’ range (Mount Lyndhurst), an overall population of 6000 individuals was estimated [@pedler1992].
Some populations were again resurveyed in 1999 [@pedler2000] and the Mount Lyndhurst population monitored again in 2007 [@pedler2007].
However, since this time the species has experienced anecdotal declines (Lyn Pedler, pers. comm. ) and has potentially contracted from some parts of its range.
Despite this, Chestnut-breasted Whiteface have had little to no research or conservation attention excluding the work of Pedler *et al.* Listed globally in the IUCN as Near Threatened, and as Rare in South Australia, the species lacks contemporary abundance and distribution data and its conservation status may reflect this uncertainty [@garnett2021].
<!-- NRS: Need to fix these lines -->

We aimed to re-assess the distribution of the Chestnut-breasted Whiteface.
We re-surveyed as many of the original 1990 surveys sites as possible, along with a selection of sites where the species had since been detected.
Through this process, we developed a standardised, repeatable survey protocol, and using this contemporary survey information we project the species future distribution under a range of emissions scenarios.
Our objective in this work was to inform the species conservation and future status assessments.
## Methods {#sec-data-methods}

### Survey Region and Site Selection

```{r}
#packages
library(dplyr)

#create a path to cbwf_project location
proj.path <- "C:/Users/admin/Documents/git/bha/cbwf_project/"

#survey data
d <- read.csv(file = paste0(proj.path, "data/table/CBWF_surveys_11052026.csv"))

n_sites <- length(unique(d$site))
tot_cbwf <- d |> 
  group_by(site) |>
  summarise(tot = sum(CBWF))
sites_w_cbwf <- nrow(subset(tot_cbwf, tot > 0))

st <- d |>
  group_by(site) |>
  summarise(st = min(time_num))

```

We conducted surveys across the known range of the Chestnut-breasted Whiteface in arid northern South Australia between June and September 2024.
Survey sites were selected to reassess historical localities and more recent records of the species, rather than to provide a random sample of available habitat.
The core set of survey sites was based on localities surveyed in [@pedler1992] including sites where birds were detected and sites where surveys were unsuccessful.
Additional sites were included from subsequent surveys in 1999–2000 [@pedler2000] and from more recent species records, particularly where these records represented areas not well covered by the 1990 survey localities.

We surveyed `r n_sites` sites in total.
These sites were distributed across four broad regions: the Coober Pedy–Glendambo region; the Oodnadatta region, including sites between Marla and Oodnadatta and between Nilpinna and William Creek; the Marree–Leigh Creek region, including the Lyndhurst–Birdsville Track area; and the Gawler Ranges–Lake Gairdner region, including sites around Mt Ive and Nonning.
Most sites occurred within arid stony plains, chenopod shrublands and associated low shrubland habitats on working pastoral properties.

We positioned surveys to cover the original historical survey locality as closely as possible, where accurate location information was available.
For some historical records, however, location information was imprecise or unavailable at the scale required to relocate the original survey point.
In these cases, we placed surveys within the approximate locality described by the historical record and in habitat judged to be broadly comparable to that originally surveyed.
A map of surveyed sites and their broad regional grouping is provided in @fig-cbwf-map, and exact coordinates of surveys sites provided i n supplementary material XXX.
<!-- add supplementary material -->

### Survey Methods

At each site, we conducted a set of repeated area searches designed to provide standardised count data while maximising the probability of detecting a rare and potentially patchily distributed species.
The standard survey unit was a 20-minute, 5-ha area search (\~224 x 224 m).
Each site comprised ten such survey cells, giving a total surveyed area of 50 ha per site.
Cells were arranged in two parallel, but spatially separated transects (\~224 x 1118 m), with five 5-ha cells in each transect (See example supplementary Fig. XX).
<!-- add supplementary material --> Transects were separated by 200m to reduce the probability of double-counting individualsindividuals.
This transect spacing additionally minimised interference between observers and playback when transects were surveyed simultaneously by different observers.
Each of the ten cells was surveyed once during a site visit.

Surveys were conducted on foot by a single observer per cell using binoculars.
Where two observers were conducting the survey, observers surveyed separate transects independently, but both beginning at the same end of their respective transects and moving sequentially through the five cells in the same direction.
This ensured a consistent direction of travel through each transect and reduced the likelihood of re-counting birds between cells, as a result of being flushed by the other observer.
In each cell, observers recorded the number of Chestnut-breasted Whiteface detected during the 20-minute search, and for each detection, observers recorded the number of individuals detected and whether the initial detection occurred before or after call playback.
To increase detectability, at the centre of each 5-ha cell, observers broadcast Chestnut-breasted Whiteface vocalisations consisting of 30 seconds of trilling calls and song, followed by a 30-second listening period, then a second 30-second broadcast followed by a second 30-second listening period.
The use of short, standardised playback bouts was intended to increase the probability of detecting birds present within the cell while limiting attraction from beyond the surveyed area.
Playback was conducted consistently across cells so that survey effort was comparable among sites.

Surveys were generally conducted in the morning (`r nrow(subset(st, st < 10.5))` of `r n_sites` sites started before 10:30AM), although the remoteness of sites and long travel distances meant that some surveys were completed later in the day.
Surveys were not conducted, where avoidable, during strong winds likely to substantially reduce detectability.
For each cell, observers recorded the date, survey start time and broad survey conditions, including wind, cloud cover, rain and temperature category.
At the centre of each cell, four habitat photographs were taken facing approximately north, east, south and west.
These photographs, together with field notes, were used to classify the broad habitat of each survey cell.

The ten-cell design provided replicated surveys within each site, increasing the probability of detecting Chestnut-breasted Whiteface where present and providing repeated count data while accounting for imperfect detection.
This design also balanced statistical and logistical requirements: greater replication was desirable because the species is rare and often occurs at low density, but survey effort had to remain feasible across a large, remote study region with widely separated sites.

### Data Analysis

#### Occupancy

We fitted a Bayesian single-season occupancy model in JAGS [@plummer2003] to estimate site-level occupancy (ψ) and detection probability (*p*).
During initial data exploration no cell-level survey condition covariates had any effect on detection, and so detection probability was held constant across all sites and visits.
The global model included five site-level covariates on occupancy (@tbl-covariates).

| Variable | Derivation | Inclusion rationale |
|------------------|------------------------|------------------------------|
| Six-month pre-survey rainfall | Cumulative precipitation in the six months prior to each site visit, extracted from Australian Gridded Climate Data (AGCD) monthly precipitation (v2-0-3) for 2024 [@evans2020] and matched to the calendar month of the site survey | Recent moisture conditions influence vegetation greenness and resource availability at the time of survey, as CBWF could be nomadic and respond to regional fluctuations in resource availability |
| Change in mean annual rainfall (1990-2024) | Slope of a pixel-wise linear regression on annual rainfall totals. Derived from AGCD monthly precipitation (v2-0-3) | Long-term drying trends may drive progressive habitat deterioration and range contraction |
| Change in days per year exceeding 40 degrees C (1990-2024) | Slope of pixel-wise linear regression multiplied by the length of the time series, expressed as absolute change in days per year. Derived from AGCD daily maximum temperature (v1-0-3) | Increasing extreme heat events are potentially associated with reduced breeding success and survival in other arid-zone passerines [@conradie2020] |
| Change in photosynthetic vegetation (PV) fractional cover (2001-2024) | Slope of a pixel-wise linear regression on monthly PV values from the CSIRO fractional cover dataset, derived from MODIS imagery [@guerschman] | Declining live cover may indicate long-term habitat degradation as a result of intensification of pastoralism [@barnetson2017; @retallack2024], affecting foraging and nesting resources |
| Change in non-photosynthetic vegetation (NPV) fractional cover (2001-2024) | Slope of a pixel-wise linear regression on monthly NPV values from the same CSIRO fractional cover dataset. | NPV is a proxy for standing dry biomass and invertebrate food web support, particularly pertinent in boom-bust arid systems where generalists may derive significant benefit from the 'brown food web' [@wijas2026; @wijas2023] |

: Occupancy covariates included in the global model. All raw data were accessed through the NCI THREDDS database. See supplementary materials XXX for detailed methods and code used to derive the covariate rasters. <!-- add supplementary materials -->

All covariates were centered and scaled to zero mean and unit variance before model fitting.
A null model with no occupancy covariates was also fitted and compared to the global model using the Watanabe-Akaike Information Criterion (WAIC), computed by marginalising the latent occupancy state analytically for each posterior sample.
WAIC is preferred over DIC for hierarchical models with discrete latent states because it averages over the full posterior rather than conditioning on a point estimate [@gelman2014].

All regression coefficients and intercept parameters were given weakly informative Normal(0, 1) priors on the logit scale.
Models were run with 3 MCMC chains of 20,000 iterations each, with a burn-in of 5,000 and thinning by a factor of 5, yielding 9,000 retained posterior samples.
Convergence was confirmed using the Gelman-Rubin diagnostic (R̂ \< 1.1 for all parameters), and model fit was assessed via a Bayesian posterior predictive check using the Freeman-Tukey statistic.

Occupancy probability was predicted spatially across the study region by applying the fitted model to 1 km² aggregated raster layers of each covariate (aggregated by a factor of 10 from the native \~100 m resolution).
The September 6-month rainfall layer was used for spatial prediction, as the majority of surveys were conducted in September.
Uncertainty was propagated by drawing 500 thinned samples from the joint posterior to produce 95% credible interval surfaces alongside the posterior mean.

#### Distribution modelling

To place the 2024 survey results in a historical and future climate context, we fitted a boosted regression tree (BRT) model of Chestnut-breasted Whiteface occurrence probability using survey records from three time periods: approximately 1960, 1990, and 2024.
The dataset combined historical presence records compiled by @pedler1992 (predominantly from the 1960s), the 1990 survey records of @pedler1992, and our 2024 survey records .
All records were coded as binary presence/absence observations.
For each record, period-matched climate data were extracted from AGCD gridded products: mean annual days with maximum temperature exceeding 40°C and mean annual rainfall (mm), averaged over a 10-year window prior to each survey period (1950–1960 for the 1960 period, 1980–1990 for the 1990 period, and 2014–2024 for the 2024 period; AGCD daily maximum temperature v1-0-3 and monthly precipitation v2-0-3).

The BRT model was fitted using the function *dismo::gbm.step* [@elith2008], with a Bernoulli error family, tree complexity of 3, learning rate of 0.001, and bag fraction of 0.5.
The optimal number of trees was determined by cross-validation within *gbm.step*.

To project occurrence probability under future climate scenarios, we obtained NARCliM2.0 regional climate model output from the NCI THREDDS server.
Bias-corrected daily maximum temperature (tasmaxAdjust) and precipitation (prAdjust) data were downloaded for five global climate models (ACCESS-ESM1-5, EC-Earth3-Veg, MPI-ESM1-2-HR, NorESM2-MM, and UKESM1-0-LL), three emissions scenarios (SSP1-2.6, SSP2-4.5, SSP3-7.0), and two weather research and forecasting climate model configurations (WRF412R3 and WRF412R5), for two 20-year periods: mid-century (2040–2059) and late-century (2080–2099).
For each GCM/SSP/RCM/period combination, we computed mean annual days exceeding 40°C and mean annual rainfall.
These were then ensemble-averaged across GCMs and RCMs within each SSP×period group, producing six projected climate scenarios.
The BRT model was applied to each scenario to generate gridded occurrence probability surfaces, which were reprojected and resampled to match the historical prediction grid for plotting in terra [@hijmans2026].

## Results

```{r}
#| label: fig-cbwf-map
#| echo: false
#| fig-cap: "2024 Chestnut-breasted Whiteface survey detections"

knitr::include_graphics(
  file.path(proj.path, "figures", "updated_figs", "CBWF_survey_outcomes.png")
)
```

A total of `r sum(tot_cbwf$tot)` Chestnut-breasted Whiteface were detected from `r sites_w_cbwf` of `r n_sites` sites (mean = `r mean(tot_cbwf$tot)`, sd = `r sd(tot_cbwf$tot)`).
Birds were only detected at sites in the Coober Pedy-Glendambo region @fig-cbwf-map

Chestnut-breasted whiteface appeared to be highly responsive to call playback.
In X% (AA of BB) of detections birds were observed coming in towards the observer to investigate call playback.
Despite this, we observed limited evidence of birds moving between cells or across transects in pursuit of call playback occurring in adjacent cells or transects, suggesting survey cell size and transect spacing was adequate to minimise double counting across surveys <!-- # PT: Maybe double check this with Darcy, but I think this was right. I certaintly didn't see much evidence that the birds followed me from one cell to the next in association with call playback. I think useful to make a commnet like this in the results as it backs up the robustness of the study design and the claims we have made about separating transects by 200 m to avoid double counting, etc. -->.

<!-- NRS: this map needs improving but will do for draft -->

<!-- PT: Agree. In particular it does not show the surveys regions, or even the Coober Pedy-Glendambo region which is suggested by the previous sentence -->

### Occupancy

```{r}
#| label: load-occu-data
occu_global  <- readRDS(paste0(proj.path, "outputs/occu_global.rds"))
occu_null    <- readRDS(paste0(proj.path, "outputs/occu_null.rds"))
waic         <- readRDS(paste0(proj.path, "outputs/waic_comparison.rds"))
delta_waic   <- round(waic$waic_null$waic - waic$waic_global$waic, 1)
```

Detection probability was estimated to be *p* = `r round(occu_global$mean$p, 2)` (95% CrI: `r round(occu_global$q2.5$p, 2)`–`r round(occu_global$q97.5$p, 2)`), indicating moderate detectability per survey cell.
Of the `r n_sites` sites surveyed, an estimated `r round(occu_global$mean$nocc, 1)` (95% CrI: `r round(occu_global$q2.5$nocc, 0)`–`r round(occu_global$q97.5$nocc, 0)`) were inferred to be occupied, showing strong alignment between model inference and field collected data (birds recorded at `r sites_w_cbwf` sites).

The global model was preferred over the null (ΔWAIC = `r delta_waic`, favouring the global model), indicating that the covariate set collectively improves out-of-sample predictive performance.
<!-- # PT: how/if at all have you done model selection here? I don't think this was described. Or have you just compared two models, the global with the null? NRS: yep just compared global to null, as we are most interested in what is driving decline, not producing the most accurate predictive model. --> Of the five occupancy covariates, however, the absolute change in extreme heat days was the only predictor whose 95% credible interval excluded zero (@fig-coeffs), with a negative effect on occupancy probability (posterior mean ≈ `r round(occu_global$mean$beta_heat, 2)` on the logit scale).
Annual rainfall, PV fractional cover, and NPV fractional cover showed positive point estimates, and 6-month rainfall a slightly negative estimate, but all four had credible intervals broadly overlapping zero, reflecting uncertainty in their individual effects given the data.

```{r}
#| label: fig-coeffs
#| fig-cap: "Posterior mean occupancy covariate effects with 95% credible intervals (logit scale). Points to the left of zero indicate a negative effect on occupancy probability."
#| out-width: 70%

knitr::include_graphics(
  file.path(proj.path, "figures", "updated_figs", "occu_global_covar_coeffs.png")
  )
```

<!-- # PT: is it possible to play with this figure a bit so it's a bit easier to read? Whole figure could be made a bit bigger, text on axes a bit bigger also. Is it possible to reduce spacing between point estimates to give more room -->

Spatial prediction indicates that the highest predicted posterior mean ψ values of Chestnut-breasted Whiteface occupancy are concentrated in the south-west of the species' range, around the Bon Bon–Glendambo area, broadly coinciding with where detection's occurred in 2024.
(@fig-psi-maps), while areas in the northern and eastern parts of the historical range where no birds were detected have substantially lower predicted occupancy probability.
The credible interval maps indicate considerable uncertainty: the lower 95% CrI surface shows that high-probability areas are restricted to the Bon Bon–Glendambo core, while the upper 95% CrI suggests potentially suitable conditions across a broader area.
Predicted occupancy mapping was only extended to the IBRA bioregions [@thackway1997] which have historical or current CBWF records, as the species is a habitat specialist of stony, low shrublands, and its occurrence at fine scales is constrained by the distribution of this habitat, which is patchy and not yet accurately mapped at the resolution of available remote sensing products.
The predicted occupancy surfaces therefore reflect broad climatic and productivity suitability across the landscape rather than realised distribution within available habitat; predicted spatial extent of occupancy should be treated as indicative of regional trends rather than literal occupied area.

```{r}
#| label: fig-psi-maps
#| fig-cap: "Spatial prediction of occupancy probability ψ: posterior mean (left), lower 95% credible interval (centre), and upper 95% credible interval (right). 2024 Survey sites are shown as black points."
#| out-width: 100%
knitr::include_graphics(
  file.path(proj.path, "figures", "updated_figs", "occupancy_psi_maps.png")
  )
```

<!-- NRS: this figure needs improvement, and the explanation of IBRA bioregions probably needs expansion in the methods? -->

### Distribution modelling

Boosted regression tree models predicted a progressive contraction of climatically suitable habitat for chestnut-breasted whiteface over the past six decades (@fig-brt-hist).
In 1960, moderate-to-high predicted occurrence probability extended across a broad swathe of the species' range.
By 1990, and more markedly by 2024, the area of high predicted occurence probability had contracted, with much of the northern and eastern range becoming less climatically suitable, driven predominantly by increasing extreme heat.

```{r}
#| label: fig-brt-hist
#| echo: false
#| fig-cap: "Boosted regression tree predicted occurrence probability of Chestnut-breasted whiteface across the study region for three historical survey periods (approximately 1960, 1990, and 2024), based on period-matched climate covariates (10 year average annual days exceeding 40°C and mean annual rainfall). Warmer colours indicate higher predicted occurrence probability. Red dots denote sites where CBWF were recorded in 2024, black dots denote all historical records/survey sites. A progressive contraction of climatically suitable habitat is evident over the past six decades, driven predominantly by increasing extreme heat."

knitr::include_graphics(
  file.path(proj.path, "figures", "updated_figs", "brt_historical_occupancy_sat.png")
)
```

Future predicted occurrence probability for Chestnut-breasted Whiteface across all three NARCliM2.0 ensemble averaged emissions scenarios suggests further reductions in climatically suitable habitat relative to the present in both mid and late century projections (@fig-brt-fut).
By mid-century (2040–2059), high-probability areas are projected to be largely confined to the south-western margin of the species current distribution.
By late century (2080–2099), suitable climate conditions are projected to contract further under all scenarios.
Under SSP1-2.6, a modest area of suitable climate persists in the south-west through the late century period, whereas under the higher emissions scenarios the majority of the species' historical range is projected to become climatically unsuitable.

```{r}
#| label: fig-brt-fut
#| echo: false
#| fig-cap: "Boosted regression tree projected occurrence probability of Chestnut-breasted Whiteface under NARCliM2.0 regional climate model ensemble scenarios for mid-century (2040–2059) and late-century (2080–2099), across three emissions pathways (SSP1-2.6, SSP2-4.5, SSP3-7.0). Warmer colours indicate higher predicted occurrence probability. Red dots denote sites where CBWF were recorded in 2024, black dots denote all historical records/survey sites. Climatically suitable habitat is projected to contract progressively relative to 2024 across all scenarios, with the most severe reductions under SSP3-7.0 by late century."

knitr::include_graphics(
  file.path(proj.path, "figures", "updated_figs", "brt_narclim_occupancy_sat.png")
)
```

<!-- # PT: I think this figure would be clearer if you had 'NARCliM2.0 emissions scenarios' written at the very top and then had the relevant scenario listed above the relevant column, and then similar on the left hand side of the figure, i.e. you had 'Future time period' written on the far left (going vertically) and then had 2040-2059 next to the top row and 2080-2099 next to the bottom row. This is as opposed to labelling each individual figure with the relevant combination of emissions scenario and time period. -->

## Discussion

Our study serves as a prime example of the critical importance of timely, systematic and scientifically robust ecological monitoring to inform species conservation.
The Chestnut-breasted Whiteface was considered an endemic, range-restricted, rare, small passerine of arid, stony, low shrublands.
It had received limited research or conservation attention in recent decades and was at high risk of declining unnoticed and undocumented.
Here, we show it's current and future projected decline to be concerning and worthy of conservation status reassessment before it's too late.

We provide the first systematic reassessment of the distribution and status of the Chestnut-breasted Whiteface across much of its known range since the surveys of Lyn Pedler and colleagues in the 1990's and 2000's [@pedler1992; @pedler2000; @pedler2007].
By revisiting historical localities, surveying more recent observation sites, and applying a repeatable site-based survey protocol, we provide an updated baseline for this species.
Our results indicate that chestnut-breasted whiteface persists in the South-West of its range around Glendambo and Coober Pedy, but was not detected from historical strongholds in the East around Mt Lyndhurst, nor from any sites along the Oodnadatta track where it has previously been recorded.
Much of these areas were predicted to have low likelihood of occupancy, under both current and future climate envelopes.
However, parts of its predicted range to the north around Indulkana in the Anangu Pitjantjatjara Yankunytjatjara lands which were not surveyed in 2024 still have a relatively high estimated occupancy probability.
Birds are regularly reported from this area by observers travelling the Stuart Highway, which is supported by our model output despite the closest survey sites recorded as unoccupied in 2024.

### Distribution has contracted and is estimated to contract further under future emissions scenarios

Of the five covariates examined, the absolute increase in days per year exceeding 40°C was the only predictor whose 95% credible interval clearly excluded zero, with a consistently negative effect on occupancy probability.
The direction and magnitude of this effect is ecologically plausible.
Small passerines in arid environments are vulnerable to extreme heat through acute lethal risks from dehydration and hyperthermia during high-temperature events [@conradie2020][@mckechnie2009].
Metabolic water requirements increase sharply above thermoneutral zone temperatures, and for species without access to free water, as is typical across much of the stony shrubland habitat used by Chestnut-breasted Whiteface, heat waves can impose lethal physiological stress even when brief [@conradie2020].
Prolonged heat can also cause chronic sublethal effects, such as progressive body mass loss, reduced nestling growth rates, and increased breeding failure in other arid zone species [@conradie2019].
The projected more-than-threefold increase in lethal dehydration risk days for Zebra Finch in some regions by the end of the 21st century [@conradie2020] demonstrates the magnitude of this threat to small arid-zone passerines under continued warming, and there is no strong reason to expect Chestnut-breasted Whiteface to be substantially more resilient.

This interpretation is reinforced by our distribution modelling, which used an expanded dataset and alternate modelling framework.
Across both historical reconstructions and future projections, change in extreme heat days was the primary axis along which climatically suitable habitat contracted.
The convergence of the two modelling approaches on the same predictor provides robust grounds for treating extreme heat as a plausible driver of the species' current distribution.
The historical pattern in the BRT projections is consistent with this: the contraction of high-probability habitat between 1960 and 2024 aligns geographically with areas that have experienced the greatest increases in extreme heat days, particularly the northern and eastern parts of the range from Oodnadatta to Mt Lyndhurst.
Future projections amplify this pattern, with a suitable climate envelope largely confined to the south-western margin of the species current distribution by mid-century under all emissions pathways, and essentially completely disappearing under higher emissions scenarios by late century.
The remaining four covariates all had credible intervals broadly spanning zero.
We expect this reflects uncertainty in the underlying data rather than these variables being ecologically unimportant.

### Occupancy is sensitive to extreme heat and concentrates in the SW of their historic distribution

Our occupancy modelling was based on a sample of sites chosen to cover historical localities.
Sites were not chosen to span the full range of environmental conditions experienced by Chestnut-breasted Whiteface.
Although this reduced our statistical power to isolate individual covariate effects, particularly where covariates were correlated, it was necessary for practical and logistical reasons and provided the greatest power to demonstrate range contraction and species decline <!-- # PT: Do you agree with this? -->.
The positive point estimates for NPV, annual rainfall and PV are nonetheless in ecologically plausible directions.
In boom-bust arid systems, rainfall pulses drive the production of standing dry biomass, which in turn supports the invertebrate food web underpinning the diet of many insectivorous passerines [@wijas2026; @wijas2023; @letnic2006].
A positive association between both PV and NPV and occupancy would be consistent with *A. pectoralis* persisting at higher density in areas with greater resources to support food webs.
Similarly, a general positive rainfall trend would be consistent with fewer extreme resource shortfalls over time.
We cannot easily explain the small negative estimate for the effect of six-month rainfall total, which may reflect a local negative relationship between recent rainfall and detection at some sites, or sampling noise.
These covariates warrant revisiting with a larger or more systematically sampled dataset to more robustly infer their individual contributions to chestnut-breasted whiteface occupancy.

*Aphelocephala pectoralis* is a habitat specialist of stony, low shrublands - a structurally distinctive vegetation type associated with gibber plains and rocky stony ground.
The fine-scale distribution of chestnut-breasted whiteface is constrained by the patchiness of this habitat within its broader range.
The climatic and productivity covariates used in the occupancy model operate at broad landscape scales, and the resulting predicted site-level occupancy surfaces reflect regional climatic and productivity suitability rather than actual habitat availability.
Because stony shrubland habitat is patchy and not yet accurately mapped at the spatial resolution of available remote sensing products, the predicted surfaces will certainly overestimate the amount of terrain within which the species could realistically occur.
Predicted occupancy probability in any given area should therefore be interpreted as indicating whether broad environmental conditions are consistent with the species' requirements, not as an estimate of occupied area.
This is a well-recognised limitation of species distribution models applied to habitat specialists with patchy microhabitat requirements <!-- # PT: do you have ref for this if you say its well-recognised limitation? -->, and resolving it would require the development of fine-scale habitat maps that capture the structural characteristics of *A. pectoralis* habitat, such as stony substrate, low open chenopod shrubland, and sparse tree cover.
Such mapping would substantially sharpen both range extent estimates and the interpretation of spatial predictions.

<!-- # PT: I ahve now included a paragraph somewhat similar to this at the start of the discussion, so I think this paragraph can potentially be deleted? -->

Taken together, the survey results, occupancy model, and BRT projections are consistent with an ongoing range contraction, with the population now apparently concentrated in the south-western core around Glendambo, Bon Bon and Coober Pedy, while formerly occupied areas in the north and east appear to have declined substantially.
The evidence presented here suggests that this contraction is at least partly driven by increasing extreme heat, a trend that is projected to continue under all plausible emissions scenarios.
These findings support considering an uplift in the conservation status of *A. pectoralis*, which is currently listed nationally as Near Threatened, a listing that may have historically reflected data deficiency [@garnett2021].

### Future directions to guide species recovery

Our study highlights several areas requiring concerted future attention: 1) The Indulkana and broader Anangu Pitjantjatjara Yankunytjatjara lands area in the north of South Australia was estimated to have a moderately high occupancy probability, and recent records from amateur ornithologists suggest the species still persists here.
However, this area was not covered in our 2024 surveys, but should be a strong focus in the near future.
2) The south-western margin of the species distribution around Glendambo and Bon Bon, which now appears to be its core range, should be urgently prioritised for ongoing ecological monitoring to estimate and track population size.
Under low emissions scenarios (SSP1-2.6), this area retains some projected climatic suitability through to the end of the century; under higher emissions, even this core area of their distribution becomes increasingly marginal.
3) The survey protocol developed here, standardised 5-ha area searches with consistent playback, replicated across ten cells per site, is suitable for direct re-use in future monitoring and provides a repeatable baseline against which future changes in detection rate and apparent occupancy can be compared.
4) Anecdotal reports have previously suggested Chestnut-breasted whiteface may be nomadic.
However, our 2024 surveys recorded the species at the exact same location as numerous historic surveys, suggesting they may in fact be highly residential.
Understanding the movement ecology of this species will be critical to informing range dynamics, population estimates, and guiding appropriate conservation actions.

### Conclusion

Ecological monitoring is frequently undervalued.
Here, we provide an example where robust, systematic ecological monitoring has identified the decline of a rare and range-restricted arid passerine, the Chestnut-breasted Whiteface.
Our work highlights the critical importance of such ecological monitoring in species conservation, and serves as a warning of what may be to come for many other arid passerines in a changing climate.

## References

::: {#refs}
:::
